\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-11 sections/chameleon.tex
Updates

\end{itemize}

@@

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed. Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression. In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.

@@

Deduction steps allow programmers to explore all the error locations one at a time (Fig. \ref{fig:deduction}). Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window. In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations to a human-readable text explanation of
their semantic connection. Programmers are free to activate any step. The active step is shown as the green button. When activating a step, some highlights switch color. The message in the explanation layer changes as well. A program in Fig. \ref{fig:deduction} generates a list of steps shown in Fig. \ref{fig:step-interface} left.

Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, they will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.

@@

Internally, deduction steps are different ways to group the error locations into two different highlighted colors. 
If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
for the Haskell programming language.
Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

@@

This has a few benefits. First, concrete types provide extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. Second, with concrete types, the debugging front end has more flexibility to change the representation of types. If the front end
understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible.

>>>>

@@

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed. Once a card is selected, the information in the conflicting types block changes according to the change of candidate expression. In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.

@@

Deduction steps allow programmers to explore all the error locations one at a time (Fig. \ref{fig:deduction}). Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window. In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations. This line is accompanied by a human-readable text explanation of their semantic connection. Programmers are free to activate any step. The active step is shown as the green button. When activating a step, some highlights switch color. The message in the explanation layer changes as well. A program in Fig. \ref{fig:deduction} generates a list of steps shown in Fig. \ref{fig:step-interface} left.

Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, programmers resolve  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, they will find a step that does not match their intention, and the type error can be fixed by modifying one of the two outlined error locations.

@@

Internally, deduction steps are different ways to group the error locations into two highlighted colors. If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
for the Haskell programming language.
Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.

@@

This has a few benefits. First, concrete types provide extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. Second, with concrete types, the debugging front end has more flexibility to change the representation of types. If the front end understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible. \tf{This sentence is not very clear}

### 2022-12-11 sections/conclusion.tex
Updates

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides accurate and comprehensive type error reports. \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

>>>>

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive type error reports. \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to interactively explore the stepwise explanations. Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages. Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

### 2022-12-11 sections/evaluation.tex
Updates

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started. 
All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to correct the code with the help of the given tool.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.

@@

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}

>>>>

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started. 
All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. They was then asked to correct the code with the help of the given tool.

\subsubsection*{\textbf{Data Collection}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.

@@

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}
